Estimating the replicability of psychology experiments after an initial failure to replicate

### 2024-08-01 manuscript/manuscript.Rmd
Moved individual project summaries to appendix and added more detail

::: {.abstract data-latex="" lang="en"}
When a replication fails, scientists have to decide whether to make a second attempt or move on. Psychology researchers who attempt to replicate studies often face this decision, given the empirical rate of replication success in psychology, which is lower than desired. Here, we report 17 re-replications of experiments for which an original replication had failed. In 5/17 of these "rescue" projects (29%), the "rescue" study mostly or fully replicated the original results, albeit with a smaller effect size; in the other 12, the second replication was also judged to have failed. We speculate that successful rescue projects were due to larger sample sizes or methodological changes such as attention checks. In the absence of obvious weaknesses in a failed replication study's sample or procedure, however, it may be most efficient to stop pursuing an effect after a single failed replication.
:::
<!-- TODO compile everything into OSF repo & link -->
<!-- TODO link individual pre-reg's in repo -->
<!-- TODO link overall pre-reg -->
<!-- TODO include in repo & link the reasons for replication failure that were offered to students -->
<!-- TODO redo/double check closeness scores for first rep and rescues  -->
<!-- TODO author contributions, acknowledgements -->

@@

<!-- Re-replication has not been studied much, so it is uncertain how often re-replications recover the results of an original study. That is, it is not well known how definitive a first replication failure is.  -->
Relatively few studies have estimated re-replication success. 
One estimate comes from @ebersole2020. After the original RP:P study reported of a replication rate around 40% [@opensciencecollaboration], @gilbert2016 raised concerns that methodological differences might explain the low replication rate. In 11 of the RP:P studies, the original authors had raised concerns about protocol fidelity prior to data collection.
Of these 11 with concerns, 10 failed to obtain significant results in the RP:P replication. 
Thus, @ebersole2020 re-replicated 10 of these 11 in larger samples, using both the RP:P protocol and a new protocol revised under advice of the original authors or other experts. This study thus tested whether larger sample sizes and/or "better" methodologies would "rescue" the failed RP:P replications.
The result was that 2 of the 10 revised protocols found a significant result (but not the 1/11 that originally was successful).
Even when a significant effect was recovered, the effect size was much diminished [@ebersole2020].
<!-- We could frame @ebersole2020 as an attempt to recover the original effects of the 9 studies with significant original effects and non-significant RP:P effects by a combination of high power and design tweaks. -->
The re-replication success rate was thus 2/9 (22%) for studies with non-significant first replications, under favorable, high-resource circumstances.

@@

A spreadsheet of projects, individual project write-ups (both first replications and rescues), individual project data and analyses for rescue projects, and the analytic code for this paper are all available at https://osf.io/cyk5w/.

Our analysis plan was pre-registered after students had selected projects, but before final data collection on the projects.
Each project was also individually pre-registered by the student conducting it.
The overall analysis is pre-registered at https://osf.io/5qz7v, individual pre-registrations are linked from https://osf.io/vkbfw.
We note one deviation from our pre-registration here: we pre-registered visual comparisons between original, first replication, and rescue projects using prediction intervals.
Prediction intervals depend on both the original effect size and variance and the variance of the comparison (replication or rescue) study [@patil2016].
Thus we cannot show a single prediction interval for the original study, but would have to show a prediction interval between each pair of studies. We thought this analysis would not offer additional clarity and hence we did not perform it.

@@

overall <- (percent_rep[1])+(1-percent_rep[1])*nrow(success)/nrow(subj)
```
## Procedure

@@

Students were required to think of reasons the first replication might not have worked, and address them if they could.
A list of possible reasons and solutions was given to students (https://osf.io/dfwk2).
In general, we encouraged students to add manipulation checks as appropriate, and to adapt their materials to online studies, which may benefit from attention checks and/or shorter duration.
For instance, the rescue of @paxton2012 switched from using the cognitive reflection task (CRT) which many online participants are over-familiar with to using the newer CRT-2 which is less overused but attempts to measure the same construct [@thomson2016].
The rescue of @jara-ettinger2022 discovered that the replication had accidentally used an illustrated version of the stimuli rather than the photographic stimuli used in the original and used the photographic stimuli instead.
@tarampi2016 had participants do a timed navigation task where they wrote L or R to indicate which way to turn at each intersection on a paper map. The replication ported this experiment online and had participants click a drop down menu to select left or right for each turn. The rescue removed the unwieldy drop down menus and opted to have participants press keys on the keyboard to indicate the direction of a turn, which seemed like a more natural interface.
Once experimental designs and analytic plans were approved by class TAs (VB and BP), students pre-registered and ran their samples.

@@

We first discuss the overall rate of replication in the rescue projects, followed by looking at the relative effect sizes of the rescues and correlates of rescue success. In the last subsection we describe the topic of each studies and focus on a few interesting case studies.

@@

Given the mix of successful and unsuccessful rescue projects, we briefly discuss all the projects and go more in depth on a few projects where we have speculations about why they turned out the way they did.

@@

@birch2007 is about a false belief task comparing ignorance and plausible knowledge conditions. The first replication, an additional class replication, and the rescue all found effects directionally consistent with the original, but much smaller.

@craig2014 had White Americans read an article about shifting racial demographics in the US, either with or without a framing that assuaged the potential threat, and found that reading the un-assuaged version led to the endorsement of more conservative positions.

@gong2019 looked at the interplay between age and willingness to donate in older adults in Hong Kong and found a three-way interaction between age, type of donation, and relationship to recipient. Neither the replication that sampled US adults nor the rescue that sampled East Asian adults recovered this interaction.

@haimovitz2016 manipulated parents' mindsets with "failure-is-enhancing" or "failure-is-debilitating" primes and looked at their reactions to a hypothetical situation in which a child received a failing grade. The first replication was directionally consistent with the original, but an additional class replication was not. The rescue added manipulation checks which passed, but did not recover the target effect.

@hopkins2016 examined whether people prefer reductive explanations to horizontal explanations. While the original found a preference for reductive explanations, the replication and rescue both did not.

@jara-ettinger2022 looked at shape bias in US adults across 3 shape exemplars and found a preference to extend nouns on the basis of shape over color or material. The replication and rescue both found an above chance preference for shape, but with widely different item-level effects and relative preferences for color and material.

@paxton2012 found that participants who completed the CRT before the moral judgement task gave more utilitarian answers than those who did the moral judgment task before the CRT. Neither the replication nor the rescue found this effect.

@porter2016 looked at how linguistic intergroup bias influences judgments of shared group membership. The replication was directionally against the original, the rescue found that effect, although the manipulation check failed. 
@schechtman2010 looked at how widely people generalize tones that previously have been paired with losses or gains, and found that the loss-paired tones were generalized more. The first replication and rescue both found the same main effect, but with a very different pattern over distance.

As discussed in methods, @tarampi2016 had participants do a timed navigation task and found an interaction between participant gender and the framing of the task (social or spatial) on performance. Neither first replication nor rescue recovered the interaction. 
@todd2016 looked at participants speed at classifying toys and guns as a function of whether they had just need a White or Black boy's face. @todd2016 found an interaction where Black faces primed faster "gun" responses and White faces faster "toy" responses. When we reached out to the original authors, they offered to run a parallel rescue on their student subject pool, which we have here as an additional replication. The first replication did not find the interaction, but both the better-powered rescue did.

@yeshurun2003 looked at the relationship between spatial attention and temporal resolution. They measured the threshold where two flashes of light are percieved as one flash versus two as a function of whether the location of the flashes was pre-cued or not. @yeshurun2003 found that cueing led to less temporal resolution. This was independently replicated by @chica2009, but was not replicated by either the student replication or the rescue.

@@

Due to the limited number of replications that are conducted and systematically reported, and the even smaller number of re-replications reported, there is a lot of uncertainty in how much replicability varies across features of the original experiment and replications.  
## Conclusion
Overall, the diminished and inconsistent effects we found in the rescues suggest that even if a re-replication "works", it may be difficult to build upon as follow-up studies will need large samples to detect small effects.

@@

<!-- Use this magic to place references here. -->

>>>>

::: {.abstract data-latex="" lang="en"}
When a replication fails, scientists have to decide whether to make a second attempt or move on.
Psychology researchers who attempt to replicate studies often face this decision, given the empirical rate of replication success in psychology, which is lower than desired.
Here, we report 17 re-replications of experiments for which an original replication had failed.
In 5/17 of these "rescue" projects (29%), the "rescue" study mostly or fully replicated the original results, albeit with a smaller effect size; in the other 12, the second replication was also judged to have failed.
We speculate that successful rescue projects were due to larger sample sizes or methodological changes such as attention checks.
In the absence of obvious weaknesses in a failed replication study's sample or procedure, however, it may be most efficient to stop pursuing an effect after a single failed replication.
:::

<!-- TODO compile everything into OSF repo & link -->

<!-- TODO link individual pre-reg's in repo -->

<!-- TODO link overall pre-reg -->

<!-- TODO include in repo & link the reasons for replication failure that were offered to students -->

<!-- TODO redo/double check closeness scores for first rep and rescues  -->

<!-- TODO author contributions, acknowledgements -->

@@

<!-- Re-replication has not been studied much, so it is uncertain how often re-replications recover the results of an original study. That is, it is not well known how definitive a first replication failure is.  -->

Relatively few studies have estimated re-replication success.
One estimate comes from @ebersole2020.
After the original RP:P study reported of a replication rate around 40% [@opensciencecollaboration], @gilbert2016 raised concerns that methodological differences might explain the low replication rate.
In 11 of the RP:P studies, the original authors had raised concerns about protocol fidelity prior to data collection.
Of these 11 with concerns, 10 failed to obtain significant results in the RP:P replication.
Thus, @ebersole2020 re-replicated 10 of these 11 in larger samples, using both the RP:P protocol and a new protocol revised under advice of the original authors or other experts.
This study thus tested whether larger sample sizes and/or "better" methodologies would "rescue" the failed RP:P replications.
The result was that 2 of the 10 revised protocols found a significant result (but not the 1/11 that originally was successful).
Even when a significant effect was recovered, the effect size was much diminished [@ebersole2020].
<!-- We could frame @ebersole2020 as an attempt to recover the original effects of the 9 studies with significant original effects and non-significant RP:P effects by a combination of high power and design tweaks. --> The re-replication success rate was thus 2/9 (22%) for studies with non-significant first replications, under favorable, high-resource circumstances.

@@

A spreadsheet of projects, individual project write-ups (both first replications and rescues), individual project data and analyses for rescue projects, and the analytic code for this paper are all available at <https://osf.io/cyk5w/>.

Our analysis plan was pre-registered after students had selected projects, but before final data collection on the projects.
Each project was also individually pre-registered by the student conducting it.
The overall analysis is pre-registered at <https://osf.io/5qz7v>, individual pre-registrations are linked from <https://osf.io/vkbfw>.
We note one deviation from our pre-registration here: we pre-registered visual comparisons between original, first replication, and rescue projects using prediction intervals.
Prediction intervals depend on both the original effect size and variance and the variance of the comparison (replication or rescue) study [@patil2016].
Thus we cannot show a single prediction interval for the original study, but would have to show a prediction interval between each pair of studies.
We thought this analysis would not offer additional clarity and hence we did not perform it.

@@

overall <- (percent_rep[1])+(1-percent_rep[1])*nrow(success)/nrow(subj)
```

## Procedure

@@

Students were required to think of reasons the first replication might not have worked, and address them if they could.
A list of possible reasons and solutions was given to students (<https://osf.io/dfwk2>).
In general, we encouraged students to add manipulation checks as appropriate, and to adapt their materials to online studies, which may benefit from attention checks and/or shorter duration.
For instance, the rescue of @paxton2012 switched from using the Cognitive Teflection Test (CRT) which many online participants are over-familiar with to using the newer CRT-2 which is less overused but attempts to measure the same construct [@thomson2016].
The rescue of @jara-ettinger2022 discovered that the replication had accidentally used an illustrated version of the stimuli rather than the photographic stimuli used in the original and used the photographic stimuli instead.
@tarampi2016 had participants do a timed navigation task where they wrote L or R to indicate which way to turn at each intersection on a paper map.
The replication ported this experiment online and had participants click a drop down menu to select left or right for each turn.
The rescue removed the unwieldy drop down menus and opted to have participants press keys on the keyboard to indicate the direction of a turn, which seemed like a more natural interface.
Once experimental designs and analytic plans were approved by class TAs (VB and BP), students pre-registered and ran their samples.

@@

We first discuss the overall rate of replication in the rescue projects, followed by looking at the relative effect sizes of the rescues and correlates of rescue success.
In the last subsection we describe several interesting case studies.

@@

Given the mix of successful and unsuccessful rescue projects, we discuss in depth a few projects where we have speculations about why they turned out the way they did.
Brief descriptions of the projects not covered in this section be found in the Appendix.

@@

@@

Due to the limited number of replications that are conducted and systematically reported, and the even smaller number of re-replications reported, there is a lot of uncertainty in how much replicability varies across features of the original experiment and replications.

## Conclusion

Overall, the diminished and inconsistent effects we found in the rescues suggest that even if a re-replication "works", it may be difficult to build upon as follow-up studies will need large samples to detect small effects.

@@

<!-- Use this magic to place references here. -->
<div id="refs"></div>

# Appendix {.unnumbered}

@birch2007 is about a false belief task comparing ignorance and plausible knowledge conditions.
The first replication and an additional class replication both found results directionally consistent with the original, but much smaller.
The rescue used a much larger sample size than both the original and the first replication and added an additional check to make sure the probabilities participants reported add up to 100%.
The rescue result was directionally consistent with the original and prior replications, but did not find a significant effect.

@craig2014 had White Americans read an article about shifting racial demographics in the US, either with or without a framing that assuaged the potential threat, and found that reading the un-assuaged version led to the endorsement of more conservative positions.
The first replication did not find a significant difference between conditions, and the rescuer speculated that this was due to a smaller sample size and a skew toward liberal political beliefs in the participant pool.
The rescue recruited the same number of participants as the original study and dropped one of the conditions to increase power.
Despite the increased power, the rescue did not replicate the original paper's finding.

@gong2019 looked at the interplay between age and willingness to donate in older adults in Hong Kong and found a three-way interaction between age, type of donation, and relationship to recipient.
The first replication sampled US adults and did not find a significant interaction.
The rescuer thought that cultural differences may have played a role and specifically sampled Asian American adults.
The rescue did not recover the interaction.

@haimovitz2016 manipulated parents' mindsets with "failure-is-enhancing" or "failure-is-debilitating" primes and looked at their reactions to a hypothetical situation in which a child received a failing grade.
The first replication was directionally inconsistent with the original, but an additional class replication was consistent.
The rescue added manipulation checks which were not present in the original study or either replications.
The manipulation checks passed, but the rescue study did not recover the target effect.

@hopkins2016 examined whether people prefer reductive explanations, which explain a result in one science with a more "fundamental" science, to horizontal explanations, which explain a result in one science in terms of the same science.
While the original study found a preference for reductive explanations, the first replication did not.
The replication's effective sample size was smaller than the original because a large proportion of participants failed the attention check.
The rescue increased made the attention check slightly easier and recruited a larger sample to account for failures, but still did not replicate the original effect.

@jara-ettinger2022 looked at shape bias in US adults across 3 shape exemplars and found a preference to extend nouns on the basis of shape over color or material.
The replication found an above-chance preference for shape, but with very high item-level variability.
The rescuer found that the first replication used stylized versions of the original stimuli taken from a different part of the paper.
The rescue used the same images of stimuli as the original, but still found strong item-level effects and relative preferences for color and material.

@paxton2012 found that participants who completed the Cognitive Reflection Test (CRT) before the moral judgement task gave more utilitarian answers than those who did the moral judgment task before the CRT.
The replication rewrote some of the CRT questions, adding a sub-condition where some of the questions featured characters with specific names.
The first replication failed to find an effect.
The rescuer removed the sub-condition and used questions from the CRT-2, out of concern that Prolific participants would be familiar with the original CRT questions.
The rescue also failed to find the effect.

@porter2016 looked at how linguistic intergroup bias influences judgments of shared group membership.
The first replication was directionally against the original result.
The rescuer obtained the direct wordings of the questions from the authors of the original paper, and noticed that they were substantially different from questions in the first replication.
The rescue used the same wordings as the original paper and replicated the effect, although the manipulation check failed.

@schechtman2010 looked at how widely people generalize tones that previously have been paired with losses or gains, and found that the loss-paired tones were generalized more.
The first replication found a statistically significant effect in the opposite direction of the original study.
It also found that people took much longer to learn which tones were associated with gains vs. losses compared to the original.
The rescuer made several changes, including asking questions about the sound quality of the participant's computer.
The original study was conducted in person while the replication and rescue were online, so differences in speaker and headphone quality might have played a role in the different result.
The replication found an effect in the same direction as the 1st replication (contrary to the original study).

As discussed in methods, @tarampi2016 had participants do a timed navigation task and found an interaction between participant gender and the framing of the task (social or spatial) on performance.
The result from the first replication was both statistically insignificant and directionally inconsistent.
The original study was conducted in person, while the replication was conducted online, which might have contributed to the inconsistency.
The rescue was also conducted online, but the rescuer made a series of user interface changes, a comprehension check, and exclusion criteria to ensure that participants understood the instructions, found the task convenient to interact with, and were engaged throughout the experiment.
Still, the rescue did not recover the interaction.

@todd2016 looked at participants speed at classifying toys and guns as a function of whether they had just need a White or Black boy's face.
@todd2016 found an interaction where Black faces primed faster "gun" responses and White faces faster "toy" responses.
When we reached out to the original authors, they offered to run a parallel rescue on their student subject pool, which we have here as an additional replication.
The first replication did not find the interaction, but both the better-powered rescue and the original authors' replication did.

@yeshurun2003 looked at the relationship between spatial attention and temporal resolution.
They measured the threshold where two flashes of light are percieved as one flash versus two as a function of whether the location of the flashes was pre-cued or not.
@yeshurun2003 found that cueing led to less temporal resolution.
This was independently replicated by @chica2009, but was not replicated by either the student replication or the rescue.
Both the replication and rescue used in-person participants because the study relies on precise timing and details of the screen which are difficult to control online.
The first replication had a smaller sample size than the original, while the rescue matched the original study's sample size.
